Astrophysical calibration of gravitational-wave detectors

### 2015-10-01 docs/cosmic_cal.tex
Finished another pass through the paper - almost there

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} and/or \ac{NSBH} systems, in addition to 
emitting detectable \acp{GW}, is also the mechanism for producing \acp{sGRB} 
\cite{1992ApJ...395L..83N}. In this scenario these events produce tightly beamed \ac{EM} emission 
parallel to the orbital angular 
momentum vector of the \ac{BNS} system. Additional evidence for the coincidence of \acp{sGRB} with
\ac{GW} events are the estimated astrophysical rates of both phenomena. Observations of \acp{sGRB} 
give a rate of XXX Mpc$^{-3}$yr$^{-1}$ which can be compared to estimates of \ac{BNS} merger rates 
of XXX-XXX Mpc$^{-3}$yr$^{-1}$, or \ac{NSBH} merger rates. The ranges of rates for \acp{BNS} 
is obtained from population synthesis models together with knowledge of the X known \ac{BNS} systems 
in our galaxy, whilst the range for \acp{NSBH} is entirely from population synthesis as not such 
systems have been observed.

% talk about the calibration idea
If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it may be possible to 
identify the host galaxy of the \ac{BNS} source.  With this information a spectroscopic redshift 
can be very accurately obtained. With knowledge of the redshift, using current best estimates for 
the Hubble constant and other cosmological parameters, the true luminosity distance can be
estimated to $\sim 1\%$ accuracy.  The \ac{GW} measurement acts as a standard siren also giving us 
a direct measurement of the luminosity distance to the source. The accuracy of such a measurement 
depends on a number of factors including the accuracy with which the \ac{GW} detector has been 
calibrated.  Hence comparison with the distance estimate from the \ac{sGRB} we can re-calibrate (or 
validate) the existing experimentally obtained calibration.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabry-P\'{e}rot Michelson interferometer; 
where 
an elaborate feedback system is used to sustain a defined measurement in arm length difference 
between the moving mirrors. A comprehensive description of the calibration procedure (in particular 
for the LIGO detectors during their fifth science run) can be found in \cite{2010NIMPA.624..223A}, 
and a brief description is giving in \cite{Vitale:2012} and references therein. For \ac{aLIGO} 
and \ac{AdV}, at the frequency range used in this project, the error in the amplitude calibration 
using the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}), which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the scale factor.}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency, but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% Standard siren basics
The idea of \ac{GW} standard sirens is directly analogous to the concept of
standard candles in \ac{EM} astronomy and was first proposed
in~\cite{1986Natur.323..310S}. Unlike the primary \ac{EM} standard candle
event, Type Ia Supernovae, the measured luminosity of a \ac{CBC} in \acp{GW} is
not only a function of distance. It also depends upon the chirp mass (a
function of the component masses) and the binary orientation with respect to
the detector network. However, measurement of the phase evolution of such an
event allows accurate determination of the chirp mass~\footnote{In reality it
is the redshifted chirp mass that is measured.}. Timing information from the
different signal arrival times at each detector in the network allows
reasonably accurate sky position determination.  Finally, amplitude variation
between differently oriented interferometers allows some (weaker) level of
determination of the binary system orientation.  A great help to this final
stage for this project is the assumption that the source has an \ac{sGRB}
counterpart.  It follows that we therefore have a well constrained prior belief
on the inclination of the source orbit with respect to the line-of-sight based
on the expected level of beaming in the \ac{sGRB} emission.

% The classic use of sirens
Standard sirens are clearly a very powerful tool for \ac{GW} cosmology since
they give us a direct measure of the absolute luminosty distance to sources.
This is distinct from Type Ia Supernovae standard candles that only provide
relative luminosty distance measures and require calibration via other methods
as part of the cosmological distance ladder. At first sight the missing
component for standard sirens is a complementary measurement of the source
redshift which would allow each event to inform the distance-redshift
relationship and hence obtain cosmological parameter estimates.  However, it
was not thought possible (until recently) that redshift measurements could be
obtained from \ac{GW} observations. In~\cite{DelPozzo:2011uf} (building on the
original idea proposed in~\cite{1986Natur.323..310S}) the idea of identifying
host galaxies within the sky position uncertainty region was developed. Each
galaxy would provide a redshift measure and would contribute to a disparate mix
of possible cosmological parameters. However, by combining multiple (${\sim}20$)
detections, the true set of cosmological parameters emerges with an expected
${\sim}5\%$ uncertainty on the Hubble constant.

% continued use of sirens
The use of \ac{GW} events with \ac{sGRB} counterparts was investigated
in~\cite{2010CQGra..27u5006S} and~\cite{2011PhRvD..83b3005Z} with respect to
the 3rd generation \ac{GW} interferometer, the \ac{ET}. It was
assumed in both studies that 1000 coincident events would be observed and
cosmological parameter estimation ability was compared with current \ac{EM}
results with focus on dark energy parameters. Other methods for cosmological
inference (see~\cite{1996PhRvD..53.2878F},~\cite{2012PhRvD..85b3535T},
and~\cite{2012PhRvD..86b3502T}) have been proposed for \ac{BNS} systems that do
not rely on any redshift measurements and instead use the distribution of
measured \acp{SNR} combined with some assumptions regarding the \ac{NS} mass
and spatial distributions. Finally we mention that for \ac{BNS} systems it has
been shown in~\cite{Messenger:2011ux} and~\cite{2013arXiv1312.1862M} that features
of the tidal and post-merger hyper-massive \ac{NS} stages of the \ac{GW}
waveform can allow the measurement of the source redshift if the \ac{NS}
equation-of-state is sufficiently well known. This development, and seperately, the method
based on \ac{SNR} distributions will allow \ac{GW} cosmological inference
without the need for complementary \ac{EM} observations.

@@

% what are sGRBs 
It is believed that \acp{sGRB} (those \acp{GRB} with duration ${<}2$ sec) are
emitted during the merger of
\ac{BNS} or \ac{NSBH} systems ~\cite{1989Natur.340..126E,1992ApJ...395L..83N}. The emission
from these events is highly beamed along the binary rotation axis and hence
only potentially observable for a fraction of the mergers. If the event
exhibits an optical afterglow then the host galaxy can be identified from which
a redshift can be obtained (e.g.~\cite{2005Natur.437..851G}).  The fraction of events with 
associated redshifts is ${\sim}1/3$. The range to which the \emph{Swift} \ac{XRT} has detected
\acp{sGRB} is $z{\sim}2$ and the nearest event with associated redshift is at
$z{\sim}0.1$~\cite{2015GCN..17278...1C} which is approximately equal to the
horizon range of the advanced \ac{GW} detector network.

% discuss joint-event rates
The predicted detection rates of \ac{BNS} events (irrespective of an \ac{sGRB}
counterpart) are derived from three main sources, population sythesis models,
extrapolations based on known galactic \ac{BNS} systems, and from observations
of \ac{sGRB} events. Various studies using the first two methods have been
compiled into an overall detection rate of $0.2$--$200$ \ac{BNS} events per
year~\cite{2013arXiv1304.0670L,2010CQGra..27q3001A}~\footnote{We should
briefly mention why the rates differ by a factor of 2.  I can't remember right
now.}. In~\cite{2012MNRAS.425.2668C} using the rate of observed \acp{sGRB} and
assuming a beaming angle of $15^{o}$ a similar event rate of $1$--$180$ is
obtained. The beaming angle is the primary factor in determining the fraction
of detected \ac{GW} \ac{BNS} signals with an \ac{EM} \ac{sGRB} counterpart and
is unfortunately highly uncertain.  Various authors have provided estimates on
the rate of joint detections to be in the range $0.02$--$7$ per
year~\cite{2013PhRvL.111r1101C,Kelley:2012fl,2014arXiv1405.2254S,2015MNRAS.448.3026W}.

@@

% the main aim of the analysis
The main aim of this work is to assess how well the calibration scale factor
defined in Eqn.~\ref{eq:scalefactor} can be estimated from a single observed
\ac{GW} associated with a particular \ac{sGRB}. As we {\it a priori} have
little knowledge of the likely location and distance of such an event we have
performed simulations of multiple events to see the expected distribution in
the accuracy of the calibration scale factor recovery.

% Describe the signal model (Taylor F2 etc...)
We use the TaylorF2 waveform approximation (see e.g.~\cite{2009PhRvD..80h4043B}
and references therein) with a 3.5 post-Newtonian expansion in phase for
modelling our signal (in both simulations and signal recovery).  For \ac{BNS}
systems we use non-spinning waveforms under the assumption that spins will be
negligible~\cm{ref?}. For \ac{NSBH} system we use a spinning waveform, but in which only
the black hole has non-negligible spin and that its rotational angular momentum
is aligned with the system's orbital angular momentum.

For a non-spinning system the general form of the frequency-domain polarisation
amplitudes are
%
\begin{align}\label{eq:signal} 
\gws_{+}(f) &\propto
\left(1+\cos^{2}\iota\right)D_{L}^{-1} \mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f,
\nu, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\ \gws_{\times}(f) &\propto
\left(\cos^{2}\iota\right)D_{L}^{-1}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \nu,
t_\mathrm{c}, \phi_{\mathrm{c}})} 
\end{align}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$,
with the symmetric mass ratio $\eta=m_{1}m_{2}/M^{2}$.  The total mass
$M=m_{1}+m_{2}$ where $m_{1}$ and $m_{2}$ are the component masses, $\iota$ is
the inclination angle, and $\nu=(\pi Mf)^{1/3}$.  The \ac{GW} strain measured
at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation} \label{eq:gravsig} 
\gws^{k}(f) = \big[ F_{+}^{k}(\alpha,
\delta, \psi)\gws_{+}(f) + F_{\times}^k(\alpha, \delta,
\psi)\gws_{\times}(f)\big] 
\end{equation}
%
where $F^{k}_{+}$ and $F^{k}_{\times}$ are the antenna response functions which
are dependent upon the polarisation angle $\psi$ and the sky position of the
source $\alpha$ and $\delta$.

% Define the detectors and the noise epoch we will be using
For this analysis we will consider the advanced generation \ac{aLIGO} and
\ac{AdV} detectors operating at their design sensitivity, with noise power
spectral densities taken from \cite{2013arXiv1304.0670L}. The detector network
we consider consists of the two \ac{aLIGO} detectors (H1 and L1) and 
the \ac{AdV} detector (V1).

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=\scf_{k}(\tilde{n}_{k}(f)+\gws_{k}(f,\btheta))
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of the true (in this case
Gaussian) strain noise, $\btheta$ is the set of waveform parameters and
$\scf_k$ is the calibration scale factor for the $k^{\rm th}$ detector, the
parameters we are interested in estimating.

@@

% mention of distance delta function prior
The primary reason why coincident observations with a \ac{sGRB} can be used as
a check of detector calibration is that the \ac{sGRB} allows us to constrain
the priors on various parameters that would normally be highly correlated with
the signal amplitude. The most important of these is that the \ac{sGRB} can
provide a very tight constraint on the source distance\footnote{For \ac{BNS}
and \ac{NSBH} systems they will be observable only in the relatively nearby
Universe ($z \sim 0.1$ and $z\sim 0.2$ for \ac{BNS} and \ac{NSBH} systems
respectively), so the relation between the measured \ac{sGRB} redshift and
distance is simple and our luminosity distance will be negligibly effected by
the choice of cosmology \MP{Check this?}.}. In this analysis we therefore
assume that the source distance is known (i.e.\ has a $\delta$-function
prior).~\cm{Should we not say a little bit more about this? A really accurate
distance from a really accurate redshift is only as good as the Hubble constant
which currently has $\sim$ few percent error?  We should also say something
about peculiar motion.}

% discuss the inclination and opening angle
Other information relevant for a coincident \ac{sGRB} is that the system is
likely to be relatively close to face-on (and therefore nearly circularly
polarised), which allows us to place prior constraints on the system
inclination angle. Beam opening angles for \acp{sGRB} are relatively poorly
constrained as they are based on only a few jet-break detections. A comparative
study of estimates and lower-limits of opening angles in 13 \acp{sGRB} given
in~\cite{2014ApJ...780..118F} provide a median opening angle value of $\sim
10^{\circ}$. We assume that the beam opening angle is a half-normal
distribution, for which the median is given by $\sqrt{2}\erf{}^{-1}(1/2)\sigma
\sim 0.67\sigma$. It follows that the median of a $\sim 10^{\circ}$ opening
angle corresponds to $\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian
prior on the inclination angle with zero mean and a standard deviation
$\sigma=14.8^{\circ}$ truncated at values below zero.

@@

% the scale factor prior
Finally, we require a prior on the calibration scale factors for each detector.
We make the assumption that in general the calibration applied to the detector
data will be correct, so want to use a prior that is peaked at $\scf_k=1$. We
also assume that the prior \ac{PDF} of the calibration scale factor is either
larger or smaller than unity by an equivalent factor, e.g.\ $p(\scf|I)=10$ is
equivelent to $p(\scf|I)=0.1$. For the scale factor in each detector we 
therefore use the following log-normal distribution as our prior
%
\begin{equation}\label{eq:scale_prior}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
%
where we chose $\sigma = 1.07$, which together with the requirement that the
mode is at unity gives $\mu = 1.15$. With these parameters the prior
probability density at 0.1 and 10 is one tenth of that at 1. Note that this
prior does not give a symmetric amount of probability about unity, but given
our simulation criterion described in Section~\ref{sec:simulations} we find
that the likelihood generally overwhelms the prior~\cm{I'm not sure about this
sentence.  Why does the symmetry matter?}. As $\scf$ is a scale factor we also
have the further constraint that it must be positive and hence the log-normal
prior is a natural choice.

@@

% a general description of the simulations 
To estimate how well the calibration scaling can be constrained for each
detector in an advanced detector network we have performed simulations with
injected \ac{BNS} and \ac{NSBH} signals. These signal injections were simulated
at a range of distances: 50\,Mpc to 500\,Mpc with 50\,Mpc increments for the
\ac{BNS} signals and 100\,Mpc to 900\,Mpc with 100\,Mpc increments for the
\ac{NSBH} signals. The simulations used the two \ac{aLIGO} detectors (H1 and
L1) and the \ac{AdV} detector (V1) and were all performed in the frequency
domain and spanned a frequency range from 20\,Hz to 400\,Hz~\cm{can you remind
me and maybe the reader of why a 400 Hz cutoff is OK?}. We assumed all
detectors were operating at their design sensitivity (as given by Figure~1 of
\cite{2013arXiv1304.0670L}). As such we added coloured Gaussian noise to each
simulation based on the sensitivity curve, but scaled with the associated
calibration scale factor.

>>>>

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} and/or \ac{NSBH} systems, in addition to 
emitting detectable \acp{GW}, is also the mechanism for producing \acp{sGRB} 
\cite{1992ApJ...395L..83N}. In this scenario these events produce tightly beamed \ac{EM} emission 
parallel to the orbital angular 
momentum vector of the \ac{BNS} system. Additional evidence for the coincidence of \acp{sGRB} with
\ac{GW} events are the estimated astrophysical rates of both phenomena. Observations of \acp{sGRB} 
give a rate of XXX\,Mpc$^{-3}$\,yr$^{-1}$ which can be compared to estimates of \ac{BNS} merger 
rates of $10^{-8}$--$10^{-5}$\,Mpc$^{-3}$\,yr$^{-1}$ \cite{2010CQGra..27q3001A}, or \ac{NSBH} merger 
rates of $6\,\times\,10^{-10}$--$1\,\times\,10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$ 
\cite{2010CQGra..27q3001A}. However, the recent work of \cite{2015MNRAS.448..928K}, which studied 
known galactic \ac{BNS} systems in more detail suggest a rate approximately five times lower. The 
estimates for \ac{NSBH} systems come entirely from population synthesis models as no such systems 
have been observed.
%The ranges of rates for \acp{BNS} is obtained from population synthesis models together with 
%knowledge of the X known \ac{BNS} systems in our galaxy, whilst the range for \acp{NSBH} is 
%entirely from population synthesis as not such systems have been observed.

If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it may be possible to 
identify the host galaxy of the \ac{BNS} source.  With this information a spectroscopic redshift 
can be very accurately obtained (see values quoted in e.g. \cite{0004-637X-664-2-1000} and 
references therein). With knowledge of the redshift, using current best 
estimates for the Hubble constant and other cosmological parameters, the true luminosity distance 
can be estimated to $\sim 1\%$ accuracy.  The \ac{GW} measurement acts as a standard siren also 
giving us a direct measurement of the luminosity distance to the source. The accuracy of such a 
measurement depends on a number of factors including the accuracy with which the \ac{GW} detector 
has been calibrated.  Hence, by comparison with the distance estimate from the \ac{sGRB} we can 
re-calibrate (or validate) the existing experimentally obtained calibration.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabry-P\'{e}rot Michelson interferometer; 
where an elaborate feedback system is used to sustain a defined measurement in arm length 
difference between the moving mirrors. A comprehensive description of the calibration procedure (in 
particular for the LIGO detectors during their fifth science run) can be found in 
\cite{2010NIMPA.624..223A}, and a brief description is giving in \cite{Vitale:2012} and references 
therein. For \ac{aLIGO} and \ac{AdV}, at the frequency range used in this project, the error in the 
amplitude calibration is roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set for the error 
estimation using the proposed method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}), which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that at a particular time we have
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the scale factor.}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency, but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% Standard siren basics
The idea of \ac{GW} standard sirens is directly analogous to the concept of
standard candles in \ac{EM} astronomy and was first proposed
in~\cite{1986Natur.323..310S}. Unlike the primary \ac{EM} standard candle
event, Type Ia Supernovae, the measured luminosity of a \ac{CBC} in \acp{GW} is
not only a function of distance. It also depends upon the chirp mass (a
function of the component masses) and the binary orientation with respect to
the detector network. However, measurement of the phase evolution of such an
event allows accurate determination of the chirp mass~\footnote{In reality it
is the redshifted chirp mass that is measured.}. Timing information from the
different signal arrival times at each detector in the network allows
reasonably accurate sky position determination.  Finally, amplitude variation
between differently oriented interferometers allows some (weaker) level of
determination of the binary system orientation.
%A great help to this final
%stage for this project is the assumption that the source has an \ac{sGRB}
%counterpart.  It follows that we therefore have a well constrained prior belief
%on the inclination of the source orbit with respect to the line-of-sight based
%on the expected level of beaming in the \ac{sGRB} emission.

% The classic use of sirens
Standard sirens are clearly a very powerful tool for \ac{GW} cosmology since
they give us a direct measure of the absolute luminosity distance to sources.
This is distinct from Type Ia Supernovae standard candles that only provide
relative luminosity distance measures and require calibration via other methods
as part of the cosmological distance ladder. At first sight the missing
component for standard sirens is a complementary measurement of the source
redshift which would allow each event to inform the distance-redshift
relationship and hence obtain cosmological parameter estimates.  However, it
was not thought possible (until recently) that redshift measurements could be
obtained from \ac{GW} observations. In~\cite{DelPozzo:2011uf} (building on the
original idea proposed in~\cite{1986Natur.323..310S}) the idea of identifying
host galaxies within the sky position uncertainty region was developed. Each
galaxy would provide a redshift measure and would contribute to a disparate mix
of possible cosmological parameters. However, by combining multiple (${\sim}20$)
detections, the true set of cosmological parameters emerges with an expected
${\sim}5\%$ uncertainty on the Hubble constant.

% continued use of sirens
The use of \ac{GW} events with \ac{sGRB} counterparts for use as a cosmological tool was 
investigated in~\cite{2010CQGra..27u5006S} and~\cite{2011PhRvD..83b3005Z} with respect to
the 3rd generation \ac{GW} interferometer, the \ac{ET}. It was
assumed in both studies that 1000 coincident events would be observed and
cosmological parameter estimation ability was compared with current \ac{EM}
results with focus on dark energy parameters. Other methods for cosmological
inference (see~\cite{1996PhRvD..53.2878F,2012PhRvD..85b3535T, 2012PhRvD..86b3502T}) have 
been proposed for \ac{BNS} systems that do
not rely on any redshift measurements and instead use the distribution of
measured \acp{SNR} combined with some assumptions regarding the \ac{NS} mass
and spatial distributions. Finally, we mention that for \ac{BNS} systems it has
been shown in~\cite{Messenger:2011ux} and~\cite{2013arXiv1312.1862M} that features
of the tidal and post-merger hyper-massive \ac{NS} stages of the \ac{GW}
waveform can allow the measurement of the source redshift if the \ac{NS}
equation-of-state is sufficiently well known. This development, and separately, the method
based on \ac{SNR} distributions will allow \ac{GW} cosmological inference
without the need for complementary \ac{EM} observations.

@@

% what are sGRBs 
It is believed that \acp{sGRB} (those \acp{GRB} with duration ${<}2$ sec) are
emitted during the merger of
\ac{BNS} or \ac{NSBH} systems ~\cite{1989Natur.340..126E,1992ApJ...395L..83N}. The emission
from these events is highly beamed along the binary rotation axis and hence
only potentially observable for a fraction of the mergers. If the event
exhibits an optical afterglow then the host galaxy can be identified from which
a redshift can be obtained (e.g.~\cite{2005Natur.437..851G}).  The fraction of events with 
associated redshifts is ${\sim}1/3$. The range to which the \emph{Swift} \ac{XRT} has detected
\acp{sGRB} is $z{\sim}2$ and the nearest event with associated redshift is at
$z{\sim}0.1$~\cite{2015GCN..17278...1C} which is approximately equal to the
horizon range of the advanced \ac{GW} detector network for \ac{BNS} systems.

% discuss joint-event rates
The predicted detection rates of \ac{BNS} events (irrespective of an \ac{sGRB}
counterpart) are derived from three main sources, population synthesis models,
extrapolations based on known galactic \ac{BNS} systems, and from observations
of \ac{sGRB} events. Various studies using the first two methods have been
compiled into an overall detection rate of $0.2$--$200$ \ac{BNS} events per
year~\cite{2013arXiv1304.0670L,2010CQGra..27q3001A}\footnote{We note that this rate is that given 
in Table 1 of \cite{2013arXiv1304.0670L} for an \ac{aLIGO} and \ac{AdV} network consisting to two 
\ac{aLIGO} detectors operating at design sensitivity. However, a higher rate of $0.4$--$400$ is 
given in \cite{2010CQGra..27q3001A}, which may differ due to the use of a third \ac{aLIGO} 
detector, but the discrepancy may also have contributions from slightly different threshold 
definitions or design sensitivity curves used. We also note the study mentioned in 
Sec.~\ref{sec:intro} that potentially lowers the expected rate by a factor of five.}. 
In~\cite{2012MNRAS.425.2668C} using the rate of observed \acp{sGRB} and
assuming a beaming angle of $15^{o}$ a similar event rate of $1$--$180$ is
obtained. The beaming angle is the primary factor in determining the fraction
of detected \ac{GW} \ac{BNS} signals with an \ac{EM} \ac{sGRB} counterpart and
is unfortunately highly uncertain.  Various authors have provided estimates on
the rate of joint detections to be in the range $0.02$--$7$ per
year~\cite{2013PhRvL.111r1101C,Kelley:2012fl,2014arXiv1405.2254S,2015MNRAS.448.3026W}.

@@

% the main aim of the analysis
The main aim of this work is to assess how well the calibration scale factor
defined in Eqn.~\ref{eq:scalefactor} can be estimated from a single observed
\ac{GW} associated with a particular \ac{sGRB}. As we {\it a priori} have
little knowledge of the likely location and distance of such an event we have
performed simulations of multiple events to see the distribution in the accuracy of the calibration 
scale factor recovery.

% Describe the signal model (Taylor F2 etc...)
For our signals we use the TaylorF2 waveform approximation (see e.g.~\cite{2009PhRvD..80h4043B}
and references therein) with a 3.5 post-Newtonian expansion in phase for
modelling our signal (in both simulations and signal recovery).  For \ac{BNS}
systems we use non-spinning waveforms under the assumption that the component spins will be small 
and have a negligible effect on the analysis\footnote{In future studies we could create our 
simulations including small spins and confirm that it does not have a significant effect on our 
results if recovery is with non-spinning templates.} (see discussion in \cite{2014ApJ...795..105S} 
and studies in \cite{2012PhRvD..86h4017B}). For \ac{NSBH} system we use a spinning waveform, but in 
which only the black hole has non-negligible spin and its rotational angular momentum is aligned 
with the system's orbital angular momentum.

For a non-spinning system the general form of the frequency-domain polarisation
amplitudes are
%
\begin{align}\label{eq:signal} 
\gws_{+}(f) &\propto
\left(1+\cos^{2}\iota\right)D_{L}^{-1} \mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f,
\mathcal{M}, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\ \gws_{\times}(f) &\propto
\left(\cos^{2}\iota\right)D_{L}^{-1}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \mathcal{M},
t_\mathrm{c}, \phi_{\mathrm{c}})} 
\end{align}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$,
with the symmetric mass ratio $\eta=m_{1}m_{2}/M^{2}$.  The total mass
$M=m_{1}+m_{2}$ where $m_{1}$ and $m_{2}$ are the component masses, $\iota$ is
the inclination angle.  The \ac{GW} strain measured
at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation} \label{eq:gravsig} 
\gws^{k}(f) = \big[ F_{+}^{k}(\alpha,
\delta, \psi)\gws_{+}(f) + F_{\times}^k(\alpha, \delta,
\psi)\gws_{\times}(f)\big] 
\end{equation}
%
where $F^{k}_{+}$ and $F^{k}_{\times}$ are the antenna response functions which
are dependent upon the polarisation angle $\psi$ and the sky position of the
source $\alpha$ and $\delta$.

% Define the detectors and the noise epoch we will be using
For this analysis we consider the advanced generation \ac{aLIGO} and
\ac{AdV} detectors operating at their design sensitivity, with noise power
spectral densities taken from \cite{2013arXiv1304.0670L}. The detector network
we consider consists of the two \ac{aLIGO} detectors (H1 and L1) and 
the \ac{AdV} detector (V1).

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=\scf_{k}(\tilde{n}_{k}(f)+\gws_{k}(f,\btheta))
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of the true (in this case
Gaussian) strain noise, $\btheta$ is the set of waveform parameters and
$\scf_k$ is the calibration scale factor for the $k^{\rm th}$ detector, which are the
parameters we are interested in estimating.

@@

% mention of distance delta function prior
The primary reason why coincident observations with a \ac{sGRB} can be used as
a check of detector calibration is that the \ac{sGRB} allows us to constrain
the priors on various parameters that would normally be highly correlated with
the signal amplitude. The most important of these is that the \ac{sGRB} can
provide a very tight constraint on the source distance. In this analysis we therefore
assume that the source distance is known (i.e.\ has a $\delta$-function
prior).~\cm{Should we not say a little bit more about this? A really accurate
distance from a really accurate redshift is only as good as the Hubble constant
which currently has $\sim$ few percent error?  We should also say something
about peculiar motion.}

% discuss the inclination and opening angle
Other information relevant for a coincident \ac{sGRB} is that the system is
likely to be relatively close to face-on (and therefore nearly circularly
polarised), which allows us to place prior constraints on the system
inclination angle. Beam opening angles for \acp{sGRB} are relatively poorly
constrained as they are based on only a few jet-break detections. A comparative
study of estimates and lower-limits of opening angles in 13 \acp{sGRB} given
in~\cite{2014ApJ...780..118F} provide a median opening angle value of $\sim
10^{\circ}$. We assume that the beam opening angle is a half-normal
distribution, for which the median is given by $\sqrt{2}\erf{}^{-1}(1/2)\sigma
\sim 0.67\sigma$. It follows that the median of a $\sim 10^{\circ}$ opening
angle corresponds to $\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian
prior on the inclination angle $\iota$ with zero mean and a standard deviation
$\sigma=14.8^{\circ}$.

@@

% the scale factor prior
Finally, we require a prior on the calibration scale factors for each detector.
We make the assumption that in general the calibration applied to the detector
data will be correct, so want to use a prior that is peaked at $\scf_k=1$. We
also assume that the prior \ac{PDF} of the calibration scale factor is either
larger or smaller than unity by an equivalent factor, e.g.\ $p(\scf|I)=10$ is
equivalent to $p(\scf|I)=0.1$. For the scale factor in each detector we 
therefore use the following log-normal distribution as our prior
%
\begin{equation}\label{eq:scale_prior}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
%
where we chose $\sigma = 1.07$, which together with the requirement that the
mode is at unity gives $\mu = 1.15$. With these parameters the prior
probability density at 0.1 and 10 is one tenth of that at 1. Note that this
prior does not give a symmetric amount of probability about unity (so out prior favour \scf\xspace 
values greater than 1), but given
our simulation criterion described in Section~\ref{sec:simulations} we find
that the likelihood generally overwhelms the prior~\cm{I'm not sure about this
sentence.  Why does the symmetry matter?}. As $\scf$ is a scale factor we also
have the further constraint that it must be positive and hence the log-normal
prior is a natural choice.

@@

% a general description of the simulations 
To estimate how well the calibration scaling can be constrained for each
detector in an advanced detector network we have performed simulations with
injected \ac{BNS} and \ac{NSBH} signals. These signal injections were simulated
at a range of distances: 50\,Mpc to 500\,Mpc with 50\,Mpc increments for the
\ac{BNS} signals and 100\,Mpc to 900\,Mpc with 100\,Mpc increments for the
\ac{NSBH} signals. The simulations used the two \ac{aLIGO} detectors (H1 and
L1) and the \ac{AdV} detector (V1) and were all performed in the frequency
domain and spanned a frequency range from 20\,Hz to 400\,Hz, which is a range that contains the 
vast majority of the \ac{SNR} and will therefore be all that is required to provide good 
constraints on \scf. We assumed all
detectors were operating at their design sensitivity (as given by Figure~1 of
\cite{2013arXiv1304.0670L}). As such we added coloured Gaussian noise to each
simulation based on the sensitivity curve, but scaled with the associated
calibration scale factor.
